---
title: "R prep. lecture notes 2023-10-25 (R Objects)"
author: "Adam Tallman"
date: "2023-10-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

###Vector concepts

There are six main ``vector concepts" to learn; atomic vector, double, integer, character, logical, and coercion.

An *atomic vector* is just a set of elements. We call it ``atomic" because its just one. You can make a vector with the function c().

```{r}
things <- c(7,6,5,4,3,2,1)
```

You can test whether something is a vector or not , with the function is.vector() and you can ask about the length of the vector using length(). We call functions like is.vector ``logical operators" because they give a true or false answer.
```{r}
is.vector(things)
length(things)
```

We can also identify positions in the vector

```{r}
things[3]
```

A *double* is a number that *can* has decimals after. We can also call this a *numeric*, which is probably the more common term. A double vector would be a vector of doubles. The vector we created above would be an example.

An *integer* is a number that cannot have decimals. You can specifically create an integer value using uppercase L. 

```{r}
3L
```
```{r}
things_2 <- c(-1L, 2L, 4L)
typeof(things_2)

```
R only saves a nuber as an integer if you include the L otherwise it'll save it as a double. 

The only reason to ever use integers as opposed to numbers is because of *floating point errors*, whch is part of all mathematics done by computers.

```{r}
sqrt(2)^2-2
```
A *character* is some text. You write a character with quotations around it.

```{r}
text <- c("Please", "write", "down", "the", "answer", "now", "everyone")
```

```{r}
typeof(text)
```
R can also do Boolean logic

```{r}
3>4
```

```{r}
4>3
```
```{r}
"class" %in% text
```

```{r}
"write" %in% text
```

You can also have a vector that is just true or false values

```{r}
things_3 <- c(TRUE, FALSE, FALSE, TRUE) 


```

Finally, note that a single vector will be saved as one of the types. If you have a single vector with a bunch of numbers, the vector will be saved as a character type.


```{r}
things_4 <- c(2,3,5,7,"thing")
```


```{r}
typeof(things_4)
```

A vector you might think is a numeric vector in fact isn't. Vectors can be coerced but only in certain directions.

```{r}
as.numeric(things_3)
as.character(things)
```
So intuitively it might make sense t get the sum of numbers of things_4, but it doesn't work

```{r}
sum(things_4)
```


We could only make things_4 a numeric vector if we removed the final value. let's find the value, with the function match(). match() takes two arguments

```{r}
things_4 <- c(2,3,5,7,"thing")

match("thing", things_4)

```

Then we can remove that value, like so

```{r}
things_4[-5]
```

We can *reassign* a value to something else.

```{r}
things_4 <- things_4[-5]
things_4
```
We still can't sum over the value yet.

```{r}
sum(things_4)
```

But now we can change thsoe numbers to actual numbers because there's no more character strings in there. 

```{r}
as.numeric(things_4)
```

That works, so we can reassign things_4 as numeric value and then sum over it

```{r}
things_4 <- as.numeric(things_4)
sum(things_4)

```

##Non-atomic vectors

Matrices are stored values in two-dimensions. The first argument gives you the number of rows and the second the number of columns

```{r}
matrix(3,3,2)
```
```{r}
matrix(c(1,2,3,4,5,6), 2,3)
```

You can change some atomic vector into a matrix with as.matrix().

```{r}
as.matrix(things)
```






















## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
